Symmetric Circuits for Rank Gates

### 2017-11-13 sections/symmetric-circuits.tex
13-Nov-2017

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background
Boolean circuits have long been as an alternative model from which to study
algorithmic complexity. A circuit is evaluated for some ordered binary input
sequence by recursively evaluating each gate in the circuit, starting from the
input gates. Suppose then we are evaluating some gate $g$ in a circuit $C$. The
procedure involves taking the input string formed by evaluating all gates input
to $g$ and applying the Boolean function that labels $g$ to that string. But
this process assumes an externally-imposed ordering on the gates input to $g$;
an assumption common in most parts of complexity theory, where the encoding of
structures for computation induces such orderings, but not in finite model
theory.

Circuit models have also been studied in finite model theory as a means for
characterising the expressive power of logics. In particular, research has
focused on symmetric circuits defined over bases of symmetric Boolean functions
(see \cite{}). The fact that the basis consist of only symmetric functions is a
crucial requirement, ensuring that the evaluation of a gate does not depend on
the order.

In this section we extend the circuit model so as to allow us to define
symmetric circuits over a broader basis of functions. We introduce the notion of
a structured function , i.e. a Boolean function whose input is indexed in
accordance with some many-sorted structure. This allows us to characterise many
natural symmetry properties of structured functions in terms of equally natural
symmetry properties of many-sorted structures. We use this to derive a general
symmetry conditions on structured functions and a symmetric circuit model whose
basis elements are symmetric structured functions.

@@

\subsection{Structured Functions and Symmetry}
Before we begin this section let us review Boolean circuits and a few useful
symmetry conditions placed on these circuits. Let $X$ be a finite set and $F:
\{0,1\}^{X} \rightarrow \{0,1\}$ be a Boolean function. We call $X$ the
\emph{index} of $F$, and denote it by $\ind(F)$. It is often the case that the
index $X$ is in some sense structured, but let us first consider the case where
$X$ is an unstructured set. In this case the only natural symmetry condition to
consider is where $F$ is invariant under the action of $\sym_n$, i.e. $F$ is a
symmetric function. Suppose instead $X$ consists of all two element subsets of
the set $V$. Here we can think of $X$ as representing all potential edges in a
graph, and $\vec{x} \in \{0,1\}^X$ as defining a graph. It is natural then to
consider the case where $F$ determines a property of the graph, i.e. where $F$
is invariant under the action of $\sym_V$ on the input (i.e renaming of the
vertices). We call such a function \emph{graph-invariant}. Suppose that $X := A
\times B$ for two non-empty sets $A$ and $B$. Here we can think of $\vec{x} \in
\{0,1\}^X$ as an (unordered) matrix, and we can consider the case where $F$ is
invariant under the action of $\sym_A \times \sym_B$. In this case we say $F$ is
\emph{matrix-invariant}. Notice that in each of these cases the symmetry
condition of interest on the function was related to the structure of the index
set.

In the rest of this subsection we develop the notion of structured function, a
Boolean function whose input is structured in accord with some structure and
whose natural symmetry conditions follow from similar conditions on the
structure.

% $X:= {n}\choose{2}$, for some $n$, we can think of $F$ as taking in some $n$
% vertex graph, and we should like $F$ to respect permutations on the graph. In
% this case we insist that $F$ is a \emph{graph symmetric} function, for all $$
% where we can think of $F$ as taking in a graph with $n$ vertices, where $X:=
% {n}\choose{2}$ as indexing a graph with $n$ vertices

% The set $X$ is often used just to provide an ordering, and so $F$ just takes
% in a binary $\vert X \vert$-tuple. While in general the order of the input
% sequence matters, it is common to consider the case where $F$ is a
% \emph{symmetric} Boolean function, i.e. for all $\sigma \in \sym_{X}$,
% $\vec{x} \in \{0,1\}^{X}$, $F(\vec{x}\cdot \sigma) = \vec{x}$. In this case
% the output of the function is independent of the ordering on the input tuple.
% Indeed, many natural Boolean functions (e.g. `and', `or', `not', `majority',
% etc.) are symmetric functions. One of the benefits of considering symmetric
% functions

Let $\tau = (R, S, \nu)$ be a many-sorted, relational signature, with $R = \{R_1
\ldots, R_t\}$ and $S = [s]$ for some $s \in \nats$, and $\nu(R_i) = (a^i_1,
\ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
[r_i]$. Let $\mathcal{D}$ be a $\tau$-structure, with the universe $D = D_1
\uplus \ldots \uplus D_s = \{(i,d) : d \in D_i\}$.

Let the \emph{index-domain} of $\mathcal{D}$ be $\rdom {\mathcal{D}} = \{(i, j)
:= i \in [r_j], R_j \text{ a relation in $\tau$} \}$. Let the \emph{index
  defined by} $\mathcal{D}$ be the set of all functions $f: \rdom(\mathcal{D})
\rightarrow D$ such that for all $R_j \in R$ and $i \in [r_j]$, $f(i, j) \in
A_{a^{j}_i}$. We abuse notation and denote the index defined by
$\ind(\mathcal{D})$. We note that, since the set of relation symbols and sort
symbols are ordered, we can also think of each $f \in \ind(\mathcal{D})$ as a
tuple in $D_{a^1_1} \times \ldots \times D_{a^1_{r_1}} \times \ldots \times
D_{a^t_1} \times D_{a^t_{r_t}}$. We will use these tuples in order to index the
inputs to a Boolean function.

For the remainder of the section fix a $\tau$-structure $\mathcal{D}$ and let $X
:= \ind (\mathcal{D})$. Let $G \leq \sym_{X}$. We note that there is a natural
action of $G$ on $X$. Furthermore, we can define an action of $G$ on functions
of the form $f: X \rightarrow H$, for any set $H$, given by $\sigma \cdot f(x) =
f(\sigma \cdot x)$ for all $x \in X$. If $H = \{0,1\}$ we say that $f$ is a
\emph{binary sequence indexed by $X$}. If $H$ is surjective, we say that
\emph{$f$ indexes $H$ by $X$} or \emph{$f$ indexes $H$ by $\mathcal{D}$}. We
note that $\aut(\mathcal{D})$ and $\iso(\mathcal{D})$ can be identified with
subgroups of $\sym_X$.

We develop a notion of equivalence from this action. Say that two functions $f,
g : X \rightarrow H$ are \emph{$G$-equivalent} iff there exists $\sigma \in G$
such that $f(x) = f(\sigma \cdot x)$ for all $x \in X$. We say $f$ and $g$ are
\emph{isomorphism-equivalent} if they are $\iso(\mathcal{D}$-equivalent. We say
that $f$ and $g$ are \emph{automorphism-equivalent} if they are $
\aut(\mathcal{D})$-equivalent. We say that $f$ and $g$ are
\emph{sort-equivalent} in the special case where every relation in $\mathcal{D}$
contains all possible tuples and $f$ and $g$ are automorphism-equivalent.

We call a function of the form $F:\{0,1\}^{X} \rightarrow \{0,1\}$ a
\emph{$\mathcal{D}$-structured function}. We note that $\ind(F) =
\ind(\mathcal{D})$, justifying our above choice of notation. We also say
\emph{the vocabulary of $F$} and \emph{the universe of $F$} to refer to the
vocabulary and universe of $\mathcal{D}$.

@@

These structured functions give us a general framework for indexing the input to
functions while also providing us with a natural set of groups from which to
develop symmetry conditions for Boolean functions. We now introduce some of
these conditions. Let $F$ be a $\mathcal{D}$-structured function. We say $F$ is
\emph{$G$-invariant} if it constant on $G$-equivalent input vectors, i.e. for
all $G$-equivalent $\vec{x},\vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) = F(\vec{y})$.
Similarly, we say $F$ is \emph{automorphism-invariant} if it is constant on
automorphism-equivalent input vectors and \emph{isomorphism-invariant} if it is
constant on isomorphism-equivalent input vectors. If all the relations in
$\mathcal{D}$ contain all possible tuples, and $F$ is automorphism-invariant,
then we say that $F$ is \emph{sort-invariant}. If the vocabulary of $F$ has two
sorts $s_1, s_2$ and one binary relation $R$ with type $(s_1, s_2)$ and $F$ is
sort-invariant, then we say $F$ is \emph{matrix-invariant}.

Notice that if $\tau$ is the empty vocabulary and $\mathcal{D}$ is a
$\tau$-structure, i.e. $\mathcal{D}$ is a pure set, then any
$\mathcal{D}$-structured function $F$ is automorphism-invariant (or
isomorphism-invariant) if, and only if, $F$ is symmetric. Also, if $\mathcal{D}$
is a graph, then $F$ is isomorphism invariant if, and only if, $F$ is
graph-invariant (except with a slight alteration, here $F$ takes in the edge
relation rather than a binary string indexed by two element sets of vertices).
Similarly, $F$ is matrix invariant (as a structured function) if, and only if,
it is matrix-invariant in the sense defined right at the beginning of this
subsection. In each of these cases choosing an appropriate structure leads to an
appropriate index and an appropriate notion of symmetry. In this way structured
functions generalise the notion of a symmetric function and allow us to capture
a broad range of structure-specific symmetry conditions in a single definition.
In the next section we will show how to define a circuit and a symmetric circuit
using bases of structured functions.

@@

\subsection{Symmetric Circuits}
We define a circuit over a basis of structured functions that are not
necessarily symmetric. As such, the circuit must include for each gate $g$ an
index on its children corresponding to the index of the function labelling $g$.

@@

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis and $\tau$ be a relational vocabulary, we define a
  \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$ computing a $q$-ary query $Q$ as a
  structure $\langle G, \Omega, \Sigma, \Lambda, L \rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ such that $\vert \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1}
    (1) \vert \leq 1$. Those gates mapped to $\tau \uplus \{0,1\}$ are called
    input gates, with those mapped to $\tau$ called relational gates and those
    mapped $\{0,1\}$ called constant gates. Those gates mapped to $\mathbb{B}$
    are called internal gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R_i})_{R_i \in
      R}$ such that $\Lambda_{R_i}$ maps each relational gate $g$ with $R_i =
    \Sigma (g)$ to the tuple $\Lambda_{R_i} (g) \in [n]^{r}$. When no ambiguity
    arises we write $\Lambda (g)$ for $\Lambda_{R_i} (g)$.
  \item $L$ associates with each internal gate $g$ a function $L(g):
    \dom(\Sigma(g)) \rightarrow G$ such that if we define a relation $W
    \subseteq G^{2}$ by $W(h_1,h_2)$ iff $h_2$ is an internal gate and $h_1$ is
    in the image of $L(h_2)$, then $(G, W)$ is a directed acyclic graph.
  \end{itemize}
\end{definition}

@@

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit. The $n$ here denotes that $C_n$ accepts structures of size $n$.
For a gate $g \in C_n$, we define the \emph{index} of $g$, denoted by $\ind(g)$,
as $\ind(\Sigma(g))$ and we let $H_g$ denote the set of all gates in the image
of $L(g)$.

For any two sets $X, Y$ we let $X^{\underline{Y}}$ denote the set of all
injective functions from $Y$ to $X$.

Let $\tau = (R_1, \ldots , R_t)$, $\mathcal{A}$ be a $\tau$-structure of size
$n$, and $\gamma: \in [n]^{\underline{U}}$. Let $\gamma \mathcal{A}$ be the
structure formed by mapping U, the universe of $\mathcal{A}$, in accordance with
$\gamma$. The evaluation of some $(\mathbb{B}, \tau)$-circuit $C$ computing a
q-ary query $Q$ proceeds by recursively evaluating the gates in the circuit. The
evaluation of the gate $g$ using $\gamma$ with input $\mathcal{A}$ is denoted by
$C[\gamma \mathcal{A}](g)$, and is given as follows
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate such that $\Sigma (g)$ is a
  $\mathcal{D}$-structured function, let $L^{\gamma}_g: \ind(g) \rightarrow
  \{0,1\}$ be defined by $L^{\gamma}_g(x) = C[\gamma \mathcal{A}](L(g)(x))$, for
  all $x \in \ind(g)$. Then $g$ evaluates to true iff $\Sigma(g) (L^{\gamma}_g)
  = 1$.
\end{enumerate}
We say that $C_n$ defines the $q$-ary query $Q \subseteq U^q$ where $\vec{a} \in
Q$ iff $C_n[\gamma \mathcal{A}](\Omega (\gamma \vec{a})) = 1$.

@@

\begin{definition}[Invariant Circuit]
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary query. We
  say $C_n$ is \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of
  size $n$, $\vec{a} \in U^{q}$, and $\gamma_1, \gamma_2: [n]^{\underline{U}}$
  we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 \vec{a})) = C[\gamma_2
  \mathcal{A}](\Omega (\gamma_2 \vec{a}))$.
\end{definition}
In general, the evaluation of a circuit depends on the presentation of the
structure (i.e. on the chosen bijection $\gamma$). If family of $(\mathcal{B},
\tau)$-circuits $\mathcal{C}$ is invariant it follows that the query computed is
a $q$-query on $\tau$-structures. It follows that if $q = 0$ then $\mathcal{C}$
computes a property of $\tau$-structures. The following lemma allows us to
recast this notion in terms of the language developed in this paper.

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. The function computed by $C_n$ is
  isomorphism-invariant iff $C_n$ is invariant.
\end{lem}

For the sake of simplicity, from now on we assume the that all circuits take in
structures over a single-sorted vocabulary.

>>>>

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background

%AD: re-wrote the introductory paragraphs to be a bit more substantive.

A Boolean circuit $C$ computing a function from $\{0,1\}^n$ to
$\{0,1\}$ is usually described as a directed acyclic graph, with each
gate $g$ that has $m$ incoming edges labelled with a Boolean function
$f_g : \{0,1\}^m \rightarrow \{0,1\}$ from a basis $\mathbb{B}$ and
exactly $n$ input gates of $C$ are labelled by variables.  If the
function $f_g$ can be arbitrary, then the circuit $C$ must impose an
order on the gates that provide the inputs to $g$.  When we say that
$C$ is a directed acyclic graph, without further ordering the nodes,
we implicitly assume that the functions in the basis $\mathbb{B}$ are
\emph{symmetric}.  That is to say $f_g$ is invariant under all
permutations of its $m$ inputs.  This unstated assumption is pervasive
in circuit complexity.  In particular, the standard Boolean basis of
AND, OR and NOT gates as well as bases with majority or threshold
gates only contain symmetric functions.

We are interested in Boolean circuits that compute queries on
structures, such as graphs.  In the case of (undirected, loop-free)
graphs on $n$ vertices, such a circuit might compute a function from
$\{0,1\}^{n \choose 2}$ to $\{0,1\}$.  In this case, the function
computed by the circuit is not necessarily invariant under all
permutations of the inputs, but it is invariant under all permutations
of the $n \choose 2$ inputs induced by permutations of $[n]$.
We are especially concerned with \emph{symmetric} circuits, that is
those where the permutations of $[n]$ extend to automorphisms of the
circuit.  Note that this use of the word ``symmetric'' is distinct to
that its use when applied to Boolean functions.  For such circuits, Anderson and Dawar~\cite{AndersonD17}
consider $\PT$-uniform circuits in the standard Boolean basis as
well one with majority gates.  It is a consequence of their results
that the latter are strictly more powerful than the former.  In
particular, they are shown to be equivalent
to the logic $\FPC$.  In contrast, we show, in this section, that adding any
further symmetric functions to the basis does not allow us to extend
the power of $\PT$-uniform symmetric circuits beyond that of $\FPC$.

As our aim is a circuit characterisation of $\FPR$, whose expressive
power is strictly greater than that of $\FPC$, we need to
consider gates corresponding to non-symmetric Boolean functions.  In
particular, we consider \emph{rank} gates whose inputs are a matrix
and whose result is invariant under row and column permutations.  To
lead up to this, we first develop a general framework of structured
Boolean functions.  These are functions whose inputs naturally encode
$\tau$-structures and where the output is invariant under the natural
symmetries of such structures.  The\emph{matrix-invariant} functions
are a natural partiruclar case.   We therefore define symmetric circuits in a
general form where gates can be labelled by \emph{isomorphism-invariant} structured
functions.

Ultimately, our aim is replicate the results Anderson and
Dawar~\cite{AndersonD17} in this more general setting.  A difficulty
we face is that the proof methods in that paper heavily rely on the
assumption that the Boolean functions computed by individual gates are
symmetric functions.  We address this difficulty in the next section.

% Boolean circuits have long acted as an alternative model from which to study
% algorithmic complexity. A circuit is evaluated for some input binary sequence by
% recursively evaluating each gate in the circuit, starting from the input gates.
% Suppose then we in the process of evaluating the circuit $C$ and are at the
% stage of evaluating the gate $g$. We take the input sequence formed from the
% evaluations of the gates input to $g$ and applying the element of the basis that
% labels $g$ to that sequence. It follows that in order to evaluate $g$ we need to
% suppose some order on the evaluations of the input gates. In many cases this
% order is simply induced by an assumed order on the whole of the circuit; an
% assumption common in most parts of complexity theory, where the encoding of
% structures for computation induces such orders, but problematic in finite model
% theory.

% Circuit models have also been studied in finite model theory, often as a means
% for characterising the expressive power of logics. In particular, research has
% focused on symmetric circuits defined over bases of symmetric Boolean functions
% (see \cite{}). The fact that the basis consist of only symmetric functions is a
% crucial requirement, ensuring that the evaluation of a gate does not depend on
% an external order. However, in this paper we will need to consider circuits with
% gates that compute rank, a function that is not symmetric.

% In this section we extend this circuit model so as to allow for the inclusion of
% non-symmetric functions in the basis. We allow each gate $g$ to index its input
% gate in accordance with the basis element labeling $g$. We introduce the notion
% of a \emph{structured function}, a Boolean function whose input is indexed in
% accordance with a many-sorted vocabulary $\tau$. The input vector to a
% structured function can then be thought of as encoding a $\tau$-structure. In
% particular, we consider structured functions that compute properties of
% $\tau$-structures, which we call \emph{isomorphism-invariant} structured
% functions. We then define what it means for a circuit over a basis of
% isomorphism-invariant structured functions to be symmetric. This generalises the
% notion of a symmetric circuit. We also show that there are limitations on the
% power of symmetric circuits over bases of symmetric functions, justifying the
% importance of this generalisation.

@@

\subsection{Structured Functions and Symmetry}
Let $X$ be a finite set and $F: \{0,1\}^{X} \rightarrow \{0,1\}$. We call $X$
the \emph{index} of $F$, and denote it by $\ind(F)$. We are often interested in
classes of functions characterised by certain symmetry conditions. Perhaps the
simplest example is the class of symmetric functions, i.e.\ those functions
invariant under the action of the symmetric group on their index. However, in
many cases this symmetry condition may be too strong, or otherwise
inappropriate. For example, suppose $X$ consists of all two element subsets of
some set $V$. We may think of $X$ as containing all potential edges in a graph,
and the input sequence $ \vec{x} \in \{0,1\}^X$ as encoding a graph. In this
case we might require instead that $F$ be invariant under the action of $\sym_V$
on $\vec{x}$. This condition guarantees that $F$ determines a graph property,
and we call such a function \emph{graph-invariant}. Another example of interest,
especially in this paper, is the case where $X := A \times B$, where $A$ and $B$
are non-empty sets, and we think of $\vec{x} \in \{0,1\}^X$ as encoding an
(unordered) matrix. It is natural then to consider the case where $F$ is
invariant under the action of $\sym_A \times \sym_B$, and we say such a function
is \emph{matrix-invariant}. In this subsection we develop a general framework
for dealing with functions that take in structures.

%AD - rewrote the paragraph to reduce the use of subscripts and superscripts.
Let $\tau = (R, S, \nu)$ be a many-sorted signature.
% with $R = \{R_1 \ldots,
% R_t\}$, $S = \{s_1, \ldots , s_p\}$, and $\nu(R_i) = (a^i_1, \ldots ,
% a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
% [r_i]$.
Let $D := \biguplus_{s \in S} D_{s} = \{(s,d) : d \in D_s\}$, be a
disjoint union of non-empty sets, indexed by $S$, i.e.\ $D$ is
potentially the universe of a $\tau$-structure. We call $D$ a \emph{$\tau$-sorted set}.
For each $P \in R$, let $P^D$ denote the trivial relation over the
many-sorted universe $D$, i.e.\ $P^D := D_{a_1} \times \ldots \times
D_{a_{r}}$, where $\nu(P) = (a_1, \ldots , a_{r)})$.   Let the \emph{index defined by the pair $(\tau, D)$} be
the set $\biguplus_{P\in R} P^{D}$.  We use this set to index the input to a Boolean
function, as such we abuse notation and denote the index defined by
$(\tau, D)$ by $\ind(\tau, D)$. We call $\tau$ the vocabulary and $D$ the universe of $\ind(\tau, D)$.

For the remainder of this subsection we fix a vocabulary $\tau$ and a
many-sorted set $D=\biguplus_{s \in S} D_{s}$
and let $X := \ind (\tau, D)$. Let $G \leq \sym_{X}$, and let
$\sortsym(D) := \prod_{s \in S} \sym_{D_{s}}$.  We call the elements of $\sortsym(D)$ \emph{sorted permutations}. Note that
$\sortsym(D)$ can be identified with a subgroup of $\sym_X$. There is an
obvious action of $G$ on $X$.  We can extend this to an action on the set of functions
of the form $f: X \rightarrow X$, for any set $H$, defined by $\sigma \cdot f(x)
= f(\sigma \cdot x)$ for all $x \in X$, $\sigma \in G$.

% If $H$ is surjective, we say that \emph{$f$ indexes $H$ by $X$} or \emph{$f$
% indexes $H$ by $\mathcal{D}$}.
We say that two functions $f, g : X \rightarrow H$ are \emph{$G$-equivalent} if
there exists $\sigma \in G$ such that $f = \sigma \cdot g$. We say $f$ and $g$
are \emph{isomorphism-equivalent} if they are $\sortsym(D)$-equivalent.
We are interested in the case where $H = \{0,1\}$, i.e.\ $f : X
\rightarrow \{0,1\}$. In this case $f$ defines for each relation $R^D_i$ a subset of that relation, and thus determines a $\tau$-structure over the universe $D$. We note that two functions $f, g: X \rightarrow
\{0,1\}$ are isomorphism-equivalent if, and only if, the two $\tau$-structures
they determine are isomorphic.

We call a function of the form $F:\{0,1\}^{X} \rightarrow \{0,1\}$ a
\emph{$(\tau, D)$-structured function}, or just a \emph{structured function}
when there is no need to emphasise the particular vocabulary.  We note that
$\ind(F) = \ind(\tau, D)$, justifying our choice of notation.  We let
\emph{the vocabulary of $F$} and \emph{the universe of $F$} denote
$\tau$ and $D$ respectively.

@@

Structured functions give us a general framework for indexing the input sequence to a Boolean function and associating with that function an appropriate symmetry condition. Let $F$ be a
$(\tau, D)$-structured function. We say $F$ is \emph{$G$-invariant} if it is
constant on $G$-equivalent input vectors, i.e.\ for all $G$-equivalent $\vec{x},
\vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) = F(\vec{y})$. We say $F$ is
\emph{isomorphism-invariant} if it is $\sortsym (D)$-invariant. Note that
a structured function with vocabulary $\tau$ is isomorphism-invariant if, and
only if, it decides a property of $\tau$-structures.

At the beginning of this subsection we defined three Boolean functions with
differently indexed inputs and corresponding symmetry conditions. We now look at
how these functions might be thought of as structured functions, and how the
symmetry conditions we discussed are naturally reproduced. Suppose $\tau$ is
single-sorted and consists of a single unary relation. Then any structured
function $F$ with vocabulary $\tau$ is isomorphism-invariant if, and only if, it
is symmetric. If instead the vocabulary of $F$ is the graph vocabulary, then $F$
is isomorphism-invariant if, and only if, $F$ decides a property of directed
graphs. Lastly, if the vocabulary of $F$ has just two sorts $(s_1, s_2)$ and one
binary relation with type $(s_1, s_2)$, then $F$ is isomorphism-invariant if,
and only if, it is matrix-invariant.

In each case a natural choice of the structure indexes the input to the
structured function appropriately and produces the appropriate symmetry
condition. It is worth mentioning that the graph-invariant case is not strictly
reproduced above, as the structured functions are rather defined over directed
graphs.

From the above observations, we may think of symmetric functions as
isomorphism-invariant structured functions over the single-sorted vocabulary with
a single unary relation. We can similarly think of matrix-invariant functions as
isomorphism-invariant structured functions over the appropriate vocabulary. More
formally, if the vocabulary of $F$ consists of $p$ sort symbols $s_1, \ldots ,
s_p$ and only one relation symbol with type $(s_1, \ldots, s_p)$ and $F$ is
isomorphism-invariant, then we say $F$ is \emph{sort-invariant}. If $p = 2$ and
$F$ is sort-invariant then we say that $F$ is \emph{matrix-invariant}.

In keeping with convention for Boolean functions, and for introducing simplicity
without a loss of generality, we assume that the universe of any structured
function is always a disjoint union of initial segments of the natural
numbers, i.e.\ if $F$ is structured function with universe $D_{s_1} \uplus \ldots
\uplus D_{s_p}$, then for all $s \in \{s_1 , \ldots , s_p\}$ there exists $d_s \in \mathbb{N}$ such
that $D_{s} = [d_s]$. We also assume that every $(\tau, D)$-structured function has the property that every $a \in D$ appears in at least one element of $\ind(\tau, D)$.

@@

\subsection{Symmetric Circuits}
% We define a circuit over a basis of structured functions that are not
% necessarily symmetric. As such, the circuit must include for each gate $g$ an
% index on its children corresponding to the index of the function labelling
% $g$.
In this section we extend the circuit definitions of Anderson and Dawar
\cite{AndersonD17}, introducing a circuit for structures that incorporates an
indexing for each gate in accordance with the Boolean function labelling that
gate. This allows us to define (and evaluate) circuits with gates labelled by
non-symmetric functions.

@@

We now define a \emph{$(\mathbb{B}, \tau)$-circuit} . This is a circuit that
computes a function on structures in the relational vocabulary $\tau$ and where
the gates are labelled by elements of the basis $\mathbb{B}$. Each element of
the basis is a $(\tau,D)$-structured function for some $\tau$ and $D$. Note
that, in the interests of keeping our definitions from getting excessively
complicated, we do not consider circuits computing functions over multi-sorted
vocabularies.
\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis and $\tau$ be a relational vocabulary, we define a
  \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$ computing a $q$-ary query $Q$ as a
  structure $\langle G, \Omega, \Sigma, \Lambda, L \rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ such that $\vert \Sigma^{-1} (0) \vert \leq 1$ and $\vert
    \Sigma^{-1} (1) \vert \leq 1$. Those gates mapped to $\tau \uplus \{0,1\}$
    are called input gates, with those mapped to $\tau$ called relational gates
    and those mapped $\{0,1\}$ called constant gates. Those gates mapped to
    $\mathbb{B}$ are called internal gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R_i})_{R_i \in
      R}$ such that $\Lambda_{R_i}$ maps each relational gate $g$ with $\Sigma
    (g) = R_i$ to the tuple $\Lambda_{R_i} (g) \in [n]^{r}$. When no ambiguity
    arises we write $\Lambda (g)$ for $\Lambda_{R_i} (g)$.
  \item $L$ associates with each internal gate $g$ a function $L(g):
    \ind(\Sigma(g)) \rightarrow G$ such that if we define a relation $W
    \subseteq G^{2}$ by $W(h_1,h_2)$ iff $h_2$ is an internal gate and $h_1$ is
    in the image of $L(h_2)$, then $(G, W)$ is a directed acyclic graph.
  \end{itemize}
\end{definition}

@@

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit. The $n$ here denotes that $C_n$ accepts structures of size $n$.
For a gate $g \in C_n$, we define the \emph{index} of $g$, denoted by $\ind(g)$,
as $\ind(\Sigma(g))$ and we let $H_g$ denote the set of all gates in the image
of $L(g)$. We also refer to the vocabulary and universe of $g$ to abbreviate
referring to the vocabulary and universe of $\Sigma(g)$.

% It is worth noting that in general computing the syntactic equivalence
% relation in polynomial time may not be possible. In fact, given two bipartite
% graphs it is possible to construct (in time polynomial in the size of the
% graphs) a circuit with gates labelled by matrix-symmetric functions such that
% two specified gates are syntacticly equivalent if,and only if, the two
% bipartite graphs are isomorphic. Since there is a polynomial time reduction
% from the graph isomorphism problem to the graph isomorphism problem on
% bipartite graphs, it follows that there is a polynomial time reduction from
% the graph isomorphism to the problem of computing syntactic equivalence. We
% formalise this result in the following proposition, and give the explicit
% construction referenced.

% \begin{prop}
%   There is a a polynomial time reduction from Graph Isomorphism to the problem
%   of deciding if two specified gates in a matrix-circuit are syntacticly
%   equivalent.
% \end{prop}
% \begin{proof}
% \end{proof}

% In contrast, in the restricted case where the gates of the circuit are
% labelled only by symmetric Boolean functions then the syntactic equivalence
% relation can be computed in polynomial time.

% \begin{prop}
%   Let $C_n$ be a circuit with symmetric gates. Then the problem of syntactic
%   equivalence relation on the gates of $C_n$ can be computed in polynomial
%   time.
% \end{prop}
% \begin{proof}
% \end{proof}

Let $\tau$ be a relational vocabulary, % = (R_1, \ldots , R_t)$
$\mathcal{A}$ a $\tau$-structure over an $n$-element universe $U$ and $\gamma
\in [n]^{\underline{U}}$.
% AD - need to define the notation [n]^{\underline{U}} in the background section
% -- DONE

Let $\gamma \mathcal{A}$ be the structure on universe $[n]$ formed by mapping
$U$ in accordance with $\gamma$. The evaluation of a $(\mathbb{B},
\tau)$-circuit $C_n$ computing a $q$-ary query $Q$ proceeds by recursively
evaluating the gates in the circuit. The evaluation of the gate $g$ for the
bijection $\gamma$ and input $\mathcal{A}$ is denoted by $C[\gamma
\mathcal{A}](g)$, and is given as follows:
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate such that $\Sigma (g)$ let $L^{\gamma}_g:
  \ind(g) \rightarrow \{0,1\}$ be defined by $L^{\gamma}_g(x) = C[\gamma
  \mathcal{A}](L(g)(x))$, for all $x \in \ind(g)$. Then $g$ evaluates to true
  iff $\Sigma(g) (L^{\gamma}_g) = 1$.
\end{enumerate}
We say that $C_n$ defines the $q$-ary query $Q \subseteq U^q$ under $\gamma$
where $\vec{a} \in Q$ iff $C_n[\gamma \mathcal{A}](\Omega (\gamma \vec{a})) =
1$.

@@

\begin{definition}[Invariant Circuit]
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary query. We
  say $C_n$ is \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of
  size $n$, $\vec{a} \in U^{q}$, and $\gamma_1, \gamma_2: [n]^{\underline{U}}$
  we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 \vec{a})) = C[\gamma_2
  \mathcal{A}](\Omega (\gamma_2 \vec{a}))$.
\end{definition}

In other words $C_n$ is invariant if the evaluation of the circuit on any
$n$-element structure $\mathcal{A}$ does not depend on the particular map
$\gamma$ that takes the universe of the structure to $[n]$. In general, the
evaluation of a circuit depends on the presentation of the structure (i.e.\ the
bijection $\gamma$). If a family of $(\mathcal{B}, \tau)$-circuits $\mathcal{C}$
is invariant it follows that the query computed is a $q$-query on
$\tau$-structures. Thus if $q = 0$ then $\mathcal{C}$ computes a property of
$\tau$-structures. The following lemma allows us to recast this notion in terms
of the language developed in this paper.

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. The function computed by $C_n$ is
  $\tau$-invariant iff $C_n$ is an invariant circuit.
\end{lem}
% AD - does this require a few lines of proof? To be added
